---
title: "DMA Coursework 2"
output: html_notebook
---
# 1. Data Preprocessing
```{r}
library(lattice)
library(ggplot2)
library(tidyr)
library(caret)
library(factoextra)
library(class)
library(dplyr)
library(ROSE)
```
## 1.1 Load Data
```{r}
raw_data=read.csv(file = "diabetes.csv",header = TRUE,sep = ",")
sapply(raw_data, summary)
sapply(raw_data,class)
```
## 1.2 Check duplicate value
```{r}
nrow(unique(raw_data))
```

## 1.3 Check if there is any missing data
```{r}
sapply(raw_data,FUN=function(x)nrow(is.na(x)))
```


## 1.4 Check the distribution of the data
```{r}
box=par(mfrow=c(1,2),mar=c(11,4,2,4))
boxplot(raw_data[,-9],las=3,main="Boxplot of All features")
boxplot(raw_data[,-c(5,9)],las=3,main="Boxplot of features without Insulin")
```

## 1.5 Handle the imbalanced
### Undersampling
```{r}
table(raw_data$Outcome)
```

### Undersampling
```{r}
data.balanced.under=ovun.sample(Outcome~.,data=raw_data,method = "under",p=0.5,seed = 106)$data
table(data.balanced.under$Outcome)
```
### Oversampling
```{r}
data.balanced.over=ovun.sample(Outcome~.,data=raw_data,method = "over",p=0.5,seed = 106)$data
table(data.balanced.over$Outcome)
```
### Both undersampling and oversampling
```{r}
data.balanced.both=ovun.sample(Outcome~.,data=raw_data,method = "both",p=0.5,seed = 106)$data
table(data.balanced.both$Outcome)
```
### Data Distribution of each dataset
```{r}
par(mfrow=c(1,4))
h1<-hist(raw_data$Outcome,breaks = 2,col = c('#69b3a2','#404080'),xlab="Label", main = "Raw Data")
text(h1$mids,h1$counts,labels=h1$counts, adj=c(0.5, -0.5))
h2<-hist(data.balanced.under$Outcome,breaks =2,col = c('#69b3a2','#404080'),xlab="Label", main = "Under-sampling")
text(h2$mids,h2$counts,labels=h2$counts, adj=c(0.5, -0.5))
h3<-hist(data.balanced.over$Outcome,breaks = 2,col = c('#69b3a2','#404080'),xlab="Label", main = "Over-sampling")
text(h3$mids,h3$counts,labels=h3$counts, adj=c(0.5, -0.5))
h4<-hist(data.balanced.both$Outcome,breaks = 2,col = c('#69b3a2','#404080'),xlab="Label", main = "Both")
text(h4$mids,h4$counts,labels=h4$counts, adj=c(0.5, -0.5))
```

## 1.6 Split Dataset
```{r}
set.seed(106)
trainInd=createDataPartition(raw_data[,8],p=0.80,list = FALSE)
training_set.raw=raw_data[trainInd,]
testing_set.raw=raw_data[-trainInd,]

trainInd=createDataPartition(data.balanced.under[,8],p=0.80,list = FALSE)
training_set.under=data.balanced.under[trainInd,]
testing_set.under=data.balanced.under[-trainInd,]

trainInd=createDataPartition(data.balanced.over[,8],p=0.80,list = FALSE)
training_set.over=data.balanced.over[trainInd,]
testing_set.over=data.balanced.over[-trainInd,]

trainInd=createDataPartition(data.balanced.both[,8],p=0.80,list = FALSE)
training_set.both=data.balanced.both[trainInd,]
testing_set.both=data.balanced.both[-trainInd,]
```
## 1.7 Standarlization
```{r}
training_set.raw[,1:8]=scale(training_set.raw[,1:8])
testing_set.raw[,1:8]=scale(testing_set.raw[,1:8])

training_set.under[,1:8]=scale(training_set.under[,1:8])
testing_set.under[,1:8]=scale(testing_set.under[,1:8])

training_set.over[,1:8]=scale(training_set.over[,1:8])
testing_set.over[,1:8]=scale(testing_set.over[,1:8])

training_set.both[,1:8]=scale(training_set.both[,1:8])
testing_set.both[,1:8]=scale(testing_set.both[,1:8])

x_train.raw=training_set.raw[,-9]
y_train.raw=training_set.raw[,9]
x_test.raw=testing_set.raw[,-9]
y_test.raw=testing_set.raw[,9]

x_train.under=training_set.under[,-9]
y_train.under=training_set.under[,9]
x_test.under=testing_set.under[,-9]
y_test.under=testing_set.under[,9]

x_train.over=training_set.over[,-9]
y_train.over=training_set.over[,9]
x_test.over=testing_set.over[,-9]
y_test.over=testing_set.over[,9]

x_train.both=training_set.both[,-9]
y_train.both=training_set.both[,9]
x_test.both=testing_set.both[,-9]
y_test.both=testing_set.both[,9]
```

## 1.8 Calculate PCA
```{r}
pca.raw=prcomp(x_train.raw)
summary(pca.raw)

pca.under=prcomp(x_train.under)
summary(pca.under)

pca.over=prcomp(x_train.over)
summary(pca.over)

pca.both=prcomp(x_train.both)
summary(pca.both)
```
```{r}
library(ggpubr)

p1=fviz_eig(pca.raw,addlabels = TRUE,ylim=c(0,40))+labs(title ="Raw Data PCA", y="% of variance")

get_eig(pca.under)
p2=fviz_eig(pca.under,addlabels = TRUE,ylim=c(0,40))+labs(title ="Under-sampling Data PCA", y="% of variance")

get_eig(pca.over)
p3=fviz_eig(pca.over,addlabels = TRUE,ylim=c(0,40))+labs(title ="Over-sampling Data PCA", y="% of variance")

get_eig(pca.both)
p4=fviz_eig(pca.both,addlabels = TRUE,ylim=c(0,40))+labs(title ='"Both" Sampling Data PCA', y="% of variance")
ggarrange(p1,p2,p3,p4,labels ="auto" ,ncol = 2,nrow = 2)
```
```{r}
x_train.raw=as.data.frame(predict(pca.raw,x_train.raw)[,1:6])
x_test.raw=as.data.frame(predict(pca.raw,x_test.raw)[,1:6])

x_train.under=as.data.frame(predict(pca.under,x_train.under)[,1:6])
x_test.under=as.data.frame(predict(pca.under,x_test.under)[,1:6])

x_train.over=as.data.frame(predict(pca.over,x_train.over)[,1:6])
x_test.over=as.data.frame(predict(pca.over,x_test.over)[,1:6])

x_train.both=as.data.frame(predict(pca.both,x_train.both)[,1:6])
x_test.both=as.data.frame(predict(pca.both,x_test.both)[,1:6])
```

# 2.KNN

```{r}
plot_knn=function(x_train,y_train,x_test,y_test){
  len=length(y_test)
  knn_evaluation=matrix(0,len,5)
  y_testInFactors=as.factor(y_test)
  for (i in seq(1:len)) {
  knn_fit=knn(train = x_train, test = x_test,cl=y_train,k=i)
  confMatrix=confusionMatrix(as.factor(knn_fit),y_testInFactors)
  knn_accuracy=as.numeric(confMatrix$overall[1])
  knn_specificity=as.numeric(confMatrix$byClass[2])
  knn_precision=as.numeric(confMatrix$byClass[5])
  knn_recall=as.numeric(confMatrix$byClass[6])
  knn_f1=as.numeric(confMatrix$byClass[7])
  knn_evaluation[i,]=c(knn_accuracy,knn_precision,knn_recall,knn_specificity,knn_f1)
  }
  knn_evaluation=as.data.frame(knn_evaluation)
  names(knn_evaluation)=c("Accuracy","Precision","Recall","Specificity","F1")
  acc=ggplot(data = knn_evaluation,aes(x=seq(1,len),y=Accuracy))+geom_line(color="red")+scale_x_continuous(name = "K Nearst Neighbor",breaks = (seq(0,len,20)))+scale_y_continuous(name = "Accuracy",labels = scales::percent)+ggtitle(paste("Accuracy with k from 1 to",collapse = " ",len))+geom_text(aes(x=which.max(Accuracy),y=max(Accuracy), label=paste(format(round(max(Accuracy)*100,2), nsmall=2),"%","at k=",which.max(Accuracy)),fontface=c("italic")),hjust=-0.1)+geom_point(aes(x=which.max(Accuracy),y=max(Accuracy),),colour="blue")
  
  pre=ggplot(data = knn_evaluation,aes(x=seq(1,len),y=Precision))+geom_line(color="red")+scale_x_continuous(name = "K Nearst Neighbor",breaks = (seq(0,len,20)))+scale_y_continuous(name = "Precision",labels = scales::percent)+ggtitle(paste("Precision with k from 1 to",collapse = " ",len))+geom_text(aes(x=which.max(Precision),y=max(Precision), label=paste(format(round(max(Precision)*100,2), nsmall=2),"%","at k=",which.max(Precision)),fontface=c("italic")),hjust=-0.1)+geom_point(aes(x=which.max(Precision),y=max(Precision),),colour="blue")
  
  if(which.max(knn_evaluation$Recall)>50){
      recall=ggplot(data = knn_evaluation,aes(x=seq(1,len),y=Recall))+geom_line(color="red")+scale_x_continuous(name = "K Nearst Neighbor",breaks = (seq(0,len,20)))+scale_y_continuous(name = "Recall",labels = scales::percent)+ggtitle(paste("Recall with k from 1 to",collapse = " ",len))+geom_text(aes(x=which.max(Recall),y=max(Recall), label=paste(format(round(max(Recall)*100,2), nsmall=2),"%","at k=",which.max(Recall)),fontface=c("italic")),hjust=1.1)+geom_point(aes(x=which.max(Recall),y=max(Recall),),colour="blue")

  }else{
      recall=ggplot(data = knn_evaluation,aes(x=seq(1,len),y=Recall))+geom_line(color="red")+scale_x_continuous(name = "K Nearst Neighbor",breaks = (seq(0,len,20)))+scale_y_continuous(name = "Recall",labels = scales::percent)+ggtitle(paste("Recall with k from 1 to",collapse = " ",len))+geom_text(aes(x=which.max(Recall),y=max(Recall), label=paste(format(round(max(Recall)*100,2), nsmall=2),"%","at k=",which.max(Recall)),fontface=c("italic")),hjust=-0.2)+geom_point(aes(x=which.max(Recall),y=max(Recall)),colour="blue")

  }
  speci=ggplot(data = knn_evaluation,aes(x=seq(1,len),y=Specificity))+geom_line(color="red")+scale_x_continuous(name = "K Nearst Neighbor",breaks = (seq(0,len,20)))
  if(which.max(knn_evaluation$F1)>50){
    f1=ggplot(data = knn_evaluation,aes(x=seq(1,len),y=F1))+geom_line(color="red")+scale_x_continuous(name = "K Nearst Neighbor",breaks = (seq(0,len,20)))+scale_y_continuous(name = "F1 Score",labels = scales::percent)+ggtitle(paste("F1 Score with k from 1 to",collapse = " ",len))+geom_text(aes(x=which.max(F1),y=max(F1), label=paste(format(round(max(F1)*100,2), nsmall=2),"%","at k=",which.max(F1)),fontface=c("italic")),hjust=1.1)+geom_point(aes(x=which.max(F1),y=max(F1)),colour="blue")
  }
  else{
    f1=ggplot(data = knn_evaluation,aes(x=seq(1,len),y=F1))+geom_line(color="red")+scale_x_continuous(name = "K Nearst Neighbor",breaks = (seq(0,len,20)))+scale_y_continuous(name = "F1 Score",labels = scales::percent)+ggtitle(paste("F1 Score with k from 1 to",collapse = " ",len))+geom_text(aes(x=which.max(F1),y=max(F1), label=paste(format(round(max(F1)*100,2), nsmall=2),"%","at k=",which.max(F1)),fontface=c("italic")),hjust=-0.1)+geom_point(aes(x=which.max(F1),y=max(F1)),colour="blue")
  }
    #Uncomment to swith output
    #return(acc);
    #return(pre)
    #return(recall)
    #return(speci)
    return(f1)
}
```

# Uncomment return(acc) in chunk 16 and comment other return instructions, rerun chunk 16 then run the following 
```{r}
p1=plot_knn(x_train = x_train.raw,y_train = y_train.raw,x_test=x_test.raw,y_test = y_test.raw)
p2=plot_knn(x_train = x_train.under,y_train = y_train.under,x_test=x_test.under,y_test = y_test.under)
p3=plot_knn(x_train = x_train.over,y_train = y_train.over,x_test=x_test.over,y_test = y_test.over)
p4=plot_knn(x_train = x_train.both,y_train = y_train.both,x_test=x_test.both,y_test = y_test.both)
ggarrange(p1,p2,p3,p4,labels =c("Raw","Under","Over","Both") ,ncol = 2,nrow = 2,hjust = -0.2)
```

# Uncomment return(pre) in chunk 16 and comment other return instructions, rerun chunk 16 then run the following 
```{r}
p1=plot_knn(x_train = x_train.raw,y_train = y_train.raw,x_test=x_test.raw,y_test = y_test.raw)
p2=plot_knn(x_train = x_train.under,y_train = y_train.under,x_test=x_test.under,y_test = y_test.under)
p3=plot_knn(x_train = x_train.over,y_train = y_train.over,x_test=x_test.over,y_test = y_test.over)
p4=plot_knn(x_train = x_train.both,y_train = y_train.both,x_test=x_test.both,y_test = y_test.both)
ggarrange(p1,p2,p3,p4,labels =c("Raw","Under","Over","Both") ,ncol = 2,nrow = 2,hjust = -0.1)
```

# Uncomment return(recall) in chunk 16 and comment other return instructions, rerun chunk 16 then run the following
```{r}
p1=plot_knn(x_train = x_train.raw,y_train = y_train.raw,x_test=x_test.raw,y_test = y_test.raw)
p2=plot_knn(x_train = x_train.under,y_train = y_train.under,x_test=x_test.under,y_test = y_test.under)
p3=plot_knn(x_train = x_train.over,y_train = y_train.over,x_test=x_test.over,y_test = y_test.over)
p4=plot_knn(x_train = x_train.both,y_train = y_train.both,x_test=x_test.both,y_test = y_test.both)
ggarrange(p1,p2,p3,p4,labels =c("Raw","Under","Over","Both") ,ncol = 2,nrow = 2,hjust = -0.1)
```

# Uncomment return(f1) in chunk 16 and comment other return instructions, rerun chunk 16 then run the following
```{r}
p1=plot_knn(x_train = x_train.raw,y_train = y_train.raw,x_test=x_test.raw,y_test = y_test.raw)
p2=plot_knn(x_train = x_train.under,y_train = y_train.under,x_test=x_test.under,y_test = y_test.under)
p3=plot_knn(x_train = x_train.over,y_train = y_train.over,x_test=x_test.over,y_test = y_test.over)
p4=plot_knn(x_train = x_train.both,y_train = y_train.both,x_test=x_test.both,y_test = y_test.both)
ggarrange(p1,p2,p3,p4,labels =c("Raw","Under","Over","Both") ,ncol = 2,nrow = 2,hjust = -0.1)
```

# 3. rPart Model
```{r}
new_trainingset.raw=x_train.raw
new_trainingset.raw$label=as.factor(y_train.raw)
new_testingset.raw=x_test.raw
new_testingset.raw$label=as.factor(y_test.raw)
new_trainingset.under=x_train.under
new_trainingset.under$label=as.factor(y_train.under)
new_testingset.under=x_test.under
new_testingset.under$label=as.factor(y_test.under)
new_trainingset.over=x_train.over
new_trainingset.over$label=as.factor(y_train.over)
new_testingset.over=x_test.over
new_testingset.over$label=as.factor(y_test.over)
new_trainingset.both=x_train.both
new_trainingset.both$label=as.factor(y_train.both)
new_testingset.both=x_test.both
new_testingset.both$label=as.factor(y_test.both)
```

```{r}
library(rpart)
rpart_mdl.raw=rpart(label~.,data=new_trainingset.raw,method='class')
plot(rpart_mdl.raw)
text(rpart_mdl.raw)
rpart_mdl.under=rpart(label~.,data=new_trainingset.under,method='class')
plot(rpart_mdl.under)
text(rpart_mdl.under)
rpart_mdl.over=rpart(label~.,data=new_trainingset.over,method='class')
plot(rpart_mdl.over)
text(rpart_mdl.over)
rpart_mdl.both=rpart(label~.,data=new_trainingset.both,method='class')
plot(rpart_mdl.both)
text(rpart_mdl.both)
```
```{r}
eval_rpart=function(y_pred,y_test){
  rpart_evaluation=matrix(0,4,2)
  conf=confusionMatrix(as.factor(y_pred),as.factor(y_test))
  acc=as.numeric(conf$overall[1])
  pre=as.numeric(conf$byClass[5])
  rec=as.numeric(conf$byClass[6])
  f1=as.numeric(conf$byClass[7])
  rpart_evaluation=cbind(rbind(acc,pre,rec,f1),c("Accuracy","Precision","Recall","F1"))
  return(rpart_evaluation)
}
```

```{r}
rpart_predict.raw=predict(rpart_mdl.raw,x_test.raw, type = 'class')
rpart_evaluation=eval_rpart(rpart_predict.raw,y_test.raw)

rpart_predict.under=predict(rpart_mdl.under,x_test.under, type = 'class')
rpart_evaluation=rbind(rpart_evaluation,eval_rpart(rpart_predict.under,y_test.under))  

rpart_predict.over=predict(rpart_mdl.over,x_test.over, type = 'class')
rpart_evaluation=rbind(rpart_evaluation,eval_rpart(rpart_predict.over,y_test.over))

rpart_predict.both=predict(rpart_mdl.both,x_test.both, type = 'class')
rpart_evaluation=rbind(rpart_evaluation,eval_rpart(rpart_predict.both,y_test.both))

rpart_evaluation=cbind(rpart_evaluation,c(rep("Raw",4),rep("Under",4),rep("Over",4),rep("Both",4)))

rpart_evaluation=as.data.frame(rpart_evaluation)
names(rpart_evaluation)=c("Value","Metric","Dataset")
rpart_evaluation$dataset=c("Raw","Under","Over","Both")
rpart_evaluation$Value=as.numeric(rpart_evaluation$Value)
ggplot(data = rpart_evaluation, aes(fill=Dataset,x=Metric,y=Value))+geom_bar(position = "dodge",stat = "identity")+scale_y_continuous(name = "Percent",labels = scales::percent)
```